Rapid long-range disynaptic inhibition explains the formation of cortical orientation maps.

### 2017-02-23 paper.tex
changed title
corrections from Evan

\title[Running Title]{Fast inhibitory loop explains functional map development with short-range inhibition.}

@@

Competitive interactions are believed to underlie variety of cortical processing, ranging from memory formation, attention and 
development of cortical functional organization (e.g. development of orientation maps in primary visual cortex). 
In the latter case, the competitive interactions happen along the cortical 
surface, with local populations of neurons reinforcing each other while competing with those displaced more distally. This 
specific configuration of lateral interactions is however in stark contrast with the known properties of the anatomical substrate, i.e.
excitatory connections (mediating reinforcement) having longer reach than inhibitory ones (mediating competition). No 
satisfactory biologically plausible resolution of this conflict between anatomical measures, and assumed cortical function has been proposed.
Recently a specific pattern of delays between different types of neurons in cat cortex has been discovered, where direct mono-synaptic
excitation has approximately the same delay, as the combined delays of the disynaptic inhibitory interactions between excitatory neurons 
(i.e. the sum of delays from excitatory to inhibitory and from inhibitory to excitatory neuron). Here we show that this specific pattern
of delays represents a biologically plausible explanation for how short-range inhibition can support competitive interactions in cortex
that underlie development of orientation maps in primary visual cortex. We demonstrate this statement analytically under simplifying conditions, 
and subsequently show using network simulations that development of orientation maps is preserved when long-range excitation, 
direct inhibitory to inhibitory interactions, and moderate inequality in the delays between excitatory and inhibitory pathways is 
added.

@@

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly range of brain functions including working memory \citep{Amit1995,Durstewitz2000},
orientation tuning \citep{Somers1995,Ben-Yishai1995}, and functional map development \citep{VonderMalsburg1973,CMVC,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, whereby local populations mutually reinforce each other via excitatory connections (short-range 
excitation) while long-range inhibition facilitates competition between the local populations and stabilizes the activity in the 
network. Such so called Mexican-hat arrangement of recurrent interactions, however, is in stark contrast to the know anatomical arrangement
of cortical circuitry: the excitatory neurons (especially in superficial layers) tend to form long-range arborizations spanning 
multiple columns while the axons of majority of inhibitory neurons are confined to area only several hundreds of micrometers 
in diameter \citep{Buzas2006,Budd2001}.

@@

A common hypothesis about how to reconcile Mexican-hat lateral interactions with anatomical reality is that the range of the effective inhibitory influence of an excitatory neuron onto other excitatory neurons (via the disynaptic pathway from excitatory to inhibitory to excitatory neurons - the disynaptic inhibition) will correspond to the cumulative reach of the direct excitatory and the direct inhibitory connections. The effective
inhibitory interactions between excitatory neurons will thus have longer range than the direct excitatory connections, supporting the use
of Mexican-hat lateral interaction in single population models (i.e. models without explicit separation of excitatory and
inhibitory populations). Even though very intuitive, this explanation has never been explicitly demonstrated, and  
it omits the fact that under the reasonable null-hypothesis of equal transmission delays on all connections, 
the recurrent disynaptic inhibition will lag the direct recurrent excitation. As it turns out this is crucial, 
as recent model analysis by Muir \etal\,\citep{Muir2014} shows that, under the assumption of uniform transmission delays, the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multi-synaptic network effects are negligible, effectively rejecting the disynaptic explanation behind Maxican-hat interaction. In conclusion, currently no satisfactory explanation of how topological functional organization develops in cortical networks that is consistent with the present anatomical findings exists.

As we will show in this study, the nature of transmission delays between different neural type (excitatory,inhibitory) combinations
holds the key to resolving this long standing open question. Recent study by Ohana \etal\,\citep{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

Here we describe the two models and their variants used in this study and finish with description of a measure
for assessing the extent to which a model orientation maps resemble their biological counterparts. Since this study heavily leans on methodology 
developed in our previous studies, we offer here only shortened description and refer readers to the original articles for details.

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelArchitecture/drawing.png}
\caption{General model architecture. (A) Inhibitory cortical population corresponds to a regular lattice of units in cortical space. (B) As in A but 
for excitatory population. (C) Thalamic neurons are modeled as simple difference-of-Gaussian filter followed by a threshold-linear transfer function. 
(D) The intra-cortical connectivity between the excitatory and inhibitory populations and their transmission delays. The intra-cortical connectivity is the only aspect of the general architecture presented here that changes between the different model variants explored in this study and each variant is detail in figure \ref{fig:model_variants}.
(F) The centers of RFs of the LGN neurons form a regular latice across the visual space covered by the model. This retinotopic mapping of connection fields between layers of the model is also maintained in the thalamo-cortical projection (lines between D and F). These thalamo-cortical projections are the only connections in the model that undergo Hebbian adaptation during the simulated visually driven development.}
\label{fig:model_architecture}
\end{figure}

@@

In sections \ref{sec:SM1} through \ref{sec:SM3} (model 1, 2 and 3; see figure \ref{fig:model_variants}) we use GCAL model \citep{Stevens2013}  which is the most advanced variant of LISSOM (Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen \etal\,\citep{CMVC}, which itself is based on earlier Self-Organizing Map models \citep{Kohonen1982}.

For clarity and consistency the following model description closely follows the methodology sections in our 
previous work \citep{Stevens2013,Antolik2011}. The architecture of the 3 GCAL variants presented in this 
paper is depicted in figure \ref{fig:model_architecture} and figure \ref{fig:model_variants}A-C. The models are implemented in the Topographica simulator, freely available at www.topographica.org. 
Each of the GCAL models consists of set of sheets of model neural units, corresponding to: (1) the photoreceptors, (2) the combined effects of the retinal ganglion cells (RGC) and ON and OFF LGN cells. Furthermore each model has one V1 sheet of units with direct excitatory and inhibitory interactions (model 1; see figure \ref{fig:model_variants}), or two 
V1 sheets, one corresponding to excitatory and one to inhibitory neurons (model 2 and 3; see figure \ref{fig:model_variants}).
The model sheet is a 2D array of computational elements (called units or, loosely, neurons), with activation and plasticity equations as 
described below and referenced by a coordinate system we will refer to as sheet coordinates, where the center of the sheet is designated (0.0,0.0).  
The number of units simulated in each sheet is determined by the density of units per unit length in both sheet dimensions. 
All cortical sheets have nominal dimensions \insval{cortex_size}$\times$\insval{cortex_size} in sheet coordinates. The sizes of the
RGC/LGN (\insval{lgn_size}$\times$\insval{lgn_size}) and photoreceptor (\insval{visual_field_size}$\times$\insval{visual_field_size}) sheets
were chosen to ensure that each unit in the receiving sheet has a complete set of connections, thus minimizing edge effects in the
RGC/LGN and V1 sheets. The density of units per $1.0\times$1.0 area is \insval{cortex_density}$\times$\insval{cortex_density} 
for the photoreceptor and RGC/LGN ON and OFF sheets, and \insval{cortex_density}$\times$\insval{cortex_density} for both cortical sheets.

\subsubsection{Simulation run-time and stimuly}

@@

The activation value $\psi_{i}$ of unit i in the photoreceptor sheet (P) is given by the gray-scale value in the chosen image at that point.
In this study we use two-dimensional elongated Gaussian patterns, who's center coordinates and orientation is sampled from uniform 
distribution that covers the area of photoreceptor sheet and full range of orientations. Each \insval{settle_steps} time steps \insval{num_gaussian_stim} Gaussian patterns are superimposed to form the visual input.

@@

The ON/OFF units are called RGC/LGN units because they represent the entire pathway between the retinal photoreceptors
and V1, including the retinal ganglion cells, LGN cells, and the connection pathways. The activation level for a unit at position $j$
in an RGC/LGN sheet at time $t$ is defined as:
%%
\begin{equation}
\eta_{j}(t)=f\left(\frac{A_{j}(t)}{c + \gamma_{L}\sum_{k}A_{k}(t)l_{kj}}\right)
\end{equation}
%%
\noindent where
%%
\begin{equation}
A_{j}(t) = \gamma_{F}\sum_{i\in F_j}\Psi_{i}(t)\omega_{ij}
\end{equation}
%%
The activation function $f$ is a half-wave rectifying function that ensures positive activation values, constant $\gamma_{F}=\insval{retina_to_lgn_strength}$ defines the overall strength of the afferent connections from the retina, constant $\gamma_{L}=\insval{lateral_lgn_strength}$ defines the strength of the lateral connections within the RGC/LGN sheet, constant $c=\insval{lateral_lgn_gain}$ defines the slope of the gain, $\Psi_{i}$ is the activation of unit $i$ taken from the set of photoreceptors
from which RGC/LGN unit $j$ receives input (its connection field $F_j$), $\omega_{ij}$ is the connection weight from unit $i$ in the
retina to unit $j$ in the RGC/LGN, and $l_{kj}$ is the lateral connection weight from RGC/LGN unit $k$ to RGC/LGN unit $j$.
Weights from the photoreceptors to units in the ON and OFF channels are set to fixed strengths with a difference-of-Gaussians
kernel ($\sigma_{center}=\insval{lgn_center_sigma}$, $\sigma_{surround}=\insval{lgn_surround_sigma}$, in sheet dimensions), with ON connection fields having a positive center and a negative surround and vice versa for OFF.  The lateral RGC/LGN weights are 2D Gaussians with kernel size $\sigma=\insval{lgn_lateral_sigma}$. The center of the afferent and lateral connection field of each ON/OFF unit is mapped to the location in the photoreceptor and lgn sheet corresponding to the location of that unit in sheet coordinates, making all these projections retinotopic.

@@

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation} \label{eq:neural_model}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
\noindent where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. The projection strength scaling factor of the afferent project $\gamma_{A}$ was set to 1.5 
based on Stevens \etal\,\citep{Stevens2013} while the values of the lateral excitatory  and inhibitory scaling factors, $\gamma_{E}$
and $\gamma_{I}$ respectively, were varied (see figure \ref{fig:figure1}) to find a balance between excitation and inhibition, and between afferent and lateral influences, to provide robust formation of activity bubbles that allows smooth maps to form.

@@

The threshold $\theta$ of all cortical excitatory units is updated at the end of each settling phase based on following 
equations:

@@

\noindent where $\Psi(t)$ is the output of the unit at time t and $\chi=\insval{activity_average_smoothing}$ is time constant 
controlling the decay of the influence of the past activities. The effect of this scaling mechanism is to bring the average activity of each 
cortical unit closer to the specified target. If the activity in a V1 unit moves away from the target during training, the threshold for activation is thus automatically raised or lowered to bring it closer to the target. Note that an alternative rule with only a single smoothing parameter (rather than
$\xi$ and  $\chi$) could be formulated, but the rule as presented here makes it simple for the modeler to set a desired target activity.

@@

Initial connection field weights are isotropic 2D Gaussians for the lateral excitatory projection and uniformly random within a Gaussian envelope for afferent and lateral inhibitory projections. Specifically, a neuron located at (i, j) will have the following weights in projection $p$:

@@

In the model, as images are presented to the photoreceptors, cortical afferent connection weights $\omega_{i,j,A}$ from the ON/OFF sheets are adjusted once per iterations (after cortical settling is completed) using a simple Hebbian learning rule. This rule results in connections that reflect correlations between the presynaptic ON/OFF unit activities and the postsynaptic cortical response. Hebbian connection weight adjustment at each iteration is dependent on the presynaptic activity, the postsynaptic response, and the Hebbian learning rate:

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelVariants/drawing.png}
\caption{The four considered intra-cortical connectivity variants. As in figure \ref{fig:model_architecture}, in all panels, the top line represent excitatory and bottom line inhibitory population of neurons. Models (A-C) are using firing rate models of neural units with instantaneous 
translation of inputs into membrane-potential (see section \ref{sec:gcal}). 
(A) Variant 1, assuming only local equaly wide excitatory and inhibitory connectivity, and ignoring direct inhibitory to inhibitory interactions. (B) Variant 2, as in A but with added direct inhibitory to inhibitory connections of equal extent as the other excitatory and inhibitory projections. (C) Variant 3, as in A but with added long-range excitatory connections represented based on Buz\'as \etal\,\citep{Buzas2006} as a second wider Gaussian. (D) The same as A but modeled with neural units that take into account membrane time-constant (see section \ref{sec:rate}).
The transmission delay of the inhibitory to excitatory projections in model 4 was varied in the experiments presented in section \ref{sec:SM4}
(figure \ref{fig:figure4}).}
\label{fig:model_variants}
\end{figure}

@@

The architecture of this model (model variant 4; figure \ref{fig:model_architecture} and \ref{fig:model_variants}D) is identical to the GCAL models 
(model variants 1 through 3; figure \ref{fig:model_variants}A-C) with the exception of the equations \ref{eq:vm_model} and \ref{eq:neural_model}, which were replaced with equation taking into account the membrane time constant:

\begin{equation}
\tau_{z} \frac{\delta\Psi_{i}}{\delta t} = - \Psi_{i} + \sum_{p}\gamma_{p}\sum_{j\in F_{ip}}f(\Psi_{j}(t-\nu_{p}))\omega_{pji} 
\end{equation}
\noindent where $\tau_{z}$ ($z \in \{E,I\}$) is the membrane time constant of excitatory and inhibitory neurons 
($\tau_{E}$ = \insval{membrane_time_constant_E} ms, $\tau_{I}$ = \insval{membrane_time_constant_I} ms), and $\nu_{p}$ is the transmission delay 
on the projection $p$ ($\nu_{EE}$ = \insval{td_EE}ms, $\nu_{EI}$ = \insval{td_EI} ms,$\nu_{IE}$ was varied (see section \ref{sec:SM4})).
As a consequence the evolution of the network dynamics have to be simulated at higher resolution. We chose the update step to be \insval{rm_update_step}ms, and we let the activity in the model to settle for \insval{rm_settle_time}ms, thus resulting in \insval{rm_settle_steps} settling steps as opposed to the \insval{settle_steps} of the GCAL model. The sub-cortical parametrization of the model is identical to the model variants 1-3 (figure \ref{fig:model_variants}A-C), but because the addition of membrane time constants substantially changes the cortical dynamics
the cortical parameters had to be re-adjusted. Specifically the strength of the thalamo-cortical connections was set to $\gamma_{A}=\insval{w_A}$,
the strength of the lateral excitatory-to-excitatory connections was set to $\gamma_{EE}=\insval{w_EE}$, the strength of the lateral excitatory-to-inhibitory connections was set to $\gamma_{EI}=\insval{w_EI}$, and the strength of the lateral inhibitory-to-excitatory connections was set to $\gamma_{IE}=\insval{w_IE}$.

@@

Model orientation maps are calculated based on the vector average method \citep{CMVC}. We first determine the preferred frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean preferred spatial frequency across all cortical neurons as the
value for the spatial frequency parameter across all subsequent analysis. Next, sine grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180$\degree$), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

\noindent and thus due to symmetry of normal distribution:

@@

\noindent Because convolution of two Normal distributions of variance $\sigma_{i}$  and $\sigma_{e}$ is normal distribution with variance $\sqrt{\sigma_{i}^2 + \sigma_{e}^2}$ and 
because $\sigma_{i} = \sigma_{e}$ we can simplify to:

@@

\item The change in point 1 will results in change in the balance of overall excitation and inhibition in the model which is critical 
for successful functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes due to the modification 1.

>>>>

\title[Running Title]{Rapid long-range disynaptic inhibition explains the formation of cortical orientation maps.}

@@

Competitive interactions are believed to underlie many types of cortical processing, ranging from memory formation, attention and 
development of cortical functional organization (e.g. development of orientation maps in primary visual cortex). 
In the latter case, the competitive interactions happen along the cortical 
surface, with local populations of neurons reinforcing each other while competing with those displaced more distally. This 
specific configuration of lateral interactions is however in stark contrast with the known properties of the anatomical substrate, i.e.
excitatory connections (mediating reinforcement) having longer reach than inhibitory ones (mediating competition). No 
satisfactory biologically plausible resolution of this conflict between anatomical measures, and assumed cortical function has been proposed.
Recently a specific pattern of delays between different types of neurons in cat cortex has been discovered, where direct mono-synaptic
excitation has approximately the same delay, as the combined delays of the disynaptic inhibitory interactions between excitatory neurons 
(i.e. the sum of delays from excitatory to inhibitory and from inhibitory to excitatory neurons). Here we show that this specific pattern
of delays represents a biologically plausible explanation for how short-range inhibition can support competitive interactions
that underlie the development of orientation maps in primary visual cortex. We demonstrate this statement analytically under simplifying conditions, 
and subsequently show using network simulations that development of orientation maps is preserved when long-range excitation, 
direct inhibitory to inhibitory interactions, and moderate inequality in the delays between excitatory and inhibitory pathways is 
added.

@@

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly a range of brain functions including working memory \citep{Amit1995,Durstewitz2000},
orientation tuning \citep{Somers1995,Ben-Yishai1995}, and functional map development \citep{VonderMalsburg1973,CMVC,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, whereby local populations mutually reinforce each other via excitatory connections (short-range 
excitation) while long-range inhibition facilitates competition between the local populations and stabilizes the activity in the 
network. Such so called Mexican-hat arrangement of recurrent interactions, however, is in stark contrast to the know anatomical arrangement
of cortical circuitry: the excitatory neurons (especially in superficial layers) tend to form long-range arborizations spanning 
multiple columns while the axons of a majority of inhibitory neurons are confined to an area only several hundreds of micrometers 
in diameter \citep{Buzas2006,Budd2001}.

@@

A common hypothesis about how to reconcile Mexican-hat lateral interactions with anatomical reality is that the range of the effective inhibitory influence of an excitatory neuron onto other excitatory neurons (via the disynaptic pathway from excitatory to inhibitory to excitatory neurons - the disynaptic inhibition) will correspond to the cumulative reach of the direct excitatory and the direct inhibitory connections. The effective
inhibitory interactions between excitatory neurons will thus have longer range than the direct excitatory connections, supporting the use
of Mexican-hat lateral interactions in population models (i.e. models without explicit separation of excitatory and
inhibitory populations). Even though very intuitive, this explanation has never been explicitly demonstrated, and  
it omits the fact that under the reasonable null-hypothesis of equal transmission delays for all connections, 
the recurrent disynaptic inhibition will lag the direct recurrent excitation. As it turns out this is crucial, 
as recent model analysis by Muir \etal\,\citep{Muir2014} shows that, under the assumption of uniform transmission delays, the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multi-synaptic network effects are negligible, effectively rejecting the disynaptic explanation behind Maxican-hat interaction. In conclusion, currently no satisfactory explanation of how topological functional organization develops in cortical networks that is consistent with the present anatomical findings exists.

As we will show in this study, the nature of the transmission delays between different neural type (excitatory,inhibitory) combinations
holds the key to resolving this long standing open question. A recent study by Ohana \etal\,\citep{Ohana2012} revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynaptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

Here we describe the two models and their variants used in this study and finish with a description of a measure
for assessing the extent to which model orientation maps resemble their biological counterparts. Since this study heavily leans on methodology 
developed in our previous studies, we offer here a short description and refer readers to the original articles for the details.

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelArchitecture/drawing.png}
\caption{General model architecture. (A) The inhibitory cortical population corresponds to a regular lattice of units in cortical space. (B) As in A but 
for the excitatory population. (C) Thalamic neurons are modeled as simple difference-of-Gaussian filters followed by a threshold-linear transfer function. 
(D) The intra-cortical connectivity between the excitatory and inhibitory populations and their transmission delays. The intra-cortical connectivity is the only aspect of the general architecture presented here that changes between the different model variants explored in this study and each variant is detailed in figure \ref{fig:model_variants}.
(F) The RF centers of the LGN neurons form a regular latice across the visual space covered by the model. This retinotopic mapping of connection fields between layers of the model is also maintained in the thalamo-cortical projection (lines between D and F). These thalamo-cortical projections are the only connections in the model that undergo Hebbian adaptation during simulated visually driven development.}
\label{fig:model_architecture}
\end{figure}

@@

In sections \ref{sec:SM1} through \ref{sec:SM3} (model 1, 2 and 3; see figure \ref{fig:model_variants}) we use the GCAL model \citep{Stevens2013}  which is the most advanced variant of the LISSOM (Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen \etal\,\citep{CMVC}, which itself is based on earlier Self-Organizing Map models \citep{Kohonen1982}.

For clarity and consistency the following model description closely follows the methodology sections in our 
previous work \citep{Stevens2013,Antolik2011}. The architecture of the 3 GCAL variants presented in this 
paper is depicted in figure \ref{fig:model_architecture} and figure \ref{fig:model_variants}A-C. The models are implemented in the Topographica simulator, freely available at www.topographica.org. 
Each of the GCAL models consists of a set of sheets of model neural units, corresponding to: (1) the photoreceptors, (2) the combined effects of the retinal ganglion cells (RGC) and ON and OFF LGN cells. Furthermore each model has one V1 sheet of units with direct excitatory and inhibitory interactions (model 1; see figure \ref{fig:model_variants}), or two 
V1 sheets, one corresponding to excitatory and one to inhibitory neurons (model 2 and 3; see figure \ref{fig:model_variants}).
The model sheet is a 2D array of computational elements (called units or, loosely, neurons), with activation and plasticity equations as 
described below and referenced by a coordinate system we will refer to as sheet coordinates, where the center of the sheet is designated (0.0,0.0).  
The number of units simulated in each sheet is determined by the density of units per unit length in both sheet dimensions. 
All cortical sheets have nominal dimensions \insval{cortex_size}$\times$\insval{cortex_size} in sheet coordinates. The sizes of the
RGC/LGN (\insval{lgn_size}$\times$\insval{lgn_size}) and photoreceptor (\insval{visual_field_size}$\times$\insval{visual_field_size}) sheets
were chosen to ensure that each unit in the receiving sheet has a complete set of connections, thus minimizing edge effects in the
RGC/LGN and V1 sheets. The density of units per $1.0\times$1.0 area is \insval{cortex_density}$\times$\insval{cortex_density} 
for the photoreceptor and RGC/LGN ON and OFF sheets, and \insval{cortex_density}$\times$\insval{cortex_density} for both cortical sheets.

\subsubsection{Simulation run-time and stimuli}

@@

The activation value $\psi_{i}$ of unit i in the photoreceptor sheet (P) is given by the gray-scale value in the chosen image at that point.
In this study we use two-dimensional elongated Gaussian patterns, whose center coordinates and orientation are sampled from a uniform 
distribution that covers the area of the photoreceptor sheet and the full range of orientations. Every \insval{settle_steps} time steps \insval{num_gaussian_stim} Gaussian patterns are superimposed to form the visual input.

@@

The ON/OFF units are called RGC/LGN units because they represent the entire pathway between the retinal photoreceptors
and V1, including the retinal ganglion cells, LGN cells, and the connection pathways. The activation level for a unit at position $j$
in an RGC/LGN sheet at time $t$ is defined as:
%%
\begin{equation}
\eta_{j}(t)=f\left(\frac{A_{j}(t)}{c + \gamma_{L}\sum_{k}A_{k}(t)l_{kj}}\right)
\end{equation}
%%
\noindent where
%%
\begin{equation}
A_{j}(t) = \gamma_{F}\sum_{i\in F_j}\Psi_{i}(t)\omega_{ij}
\end{equation}
%%
The activation function $f$ is a half-wave rectifying function that ensures positive activation values, constant $\gamma_{F}=\insval{retina_to_lgn_strength}$ defines the overall strength of the afferent connections from the retina, constant $\gamma_{L}=\insval{lateral_lgn_strength}$ defines the strength of the lateral connections within the RGC/LGN sheet, constant $c=\insval{lateral_lgn_gain}$ defines the slope of the gain, $\Psi_{i}$ is the activation of unit $i$ taken from the set of photoreceptors
from which RGC/LGN unit $j$ receives input (its connection field $F_j$), $\omega_{ij}$ is the connection weight from unit $i$ in the
retina to unit $j$ in the RGC/LGN, and $l_{kj}$ is the lateral connection weight from RGC/LGN unit $k$ to RGC/LGN unit $j$.
Weights from the photoreceptors to units in the ON and OFF channels are set to fixed strengths with a difference-of-Gaussians
kernel ($\sigma_{center}=\insval{lgn_center_sigma}$, $\sigma_{surround}=\insval{lgn_surround_sigma}$, in sheet dimensions), with ON connection fields having a positive center and a negative surround and vice versa for OFF.  The lateral RGC/LGN weights are 2D Gaussians with kernel size $\sigma=\insval{lgn_lateral_sigma}$. The center of the afferent and lateral connection field of each ON/OFF unit is mapped to the location in the photoreceptor and LGN sheet corresponding to the location of that unit in sheet coordinates, making all these projections retinotopic.

@@

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation} \label{eq:neural_model}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
\noindent where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. The projection strength scaling factor of the afferent projection $\gamma_{A}$ was set to 1.5 
based on Stevens \etal\,\citep{Stevens2013} while the values of the lateral excitatory and inhibitory scaling factors, $\gamma_{E}$
and $\gamma_{I}$ respectively, were varied (see figure \ref{fig:figure1}) to find a balance between excitation and inhibition, and between afferent and lateral influences, to provide robust formation of activity bubbles that facilitates the formation of smooth maps.

@@

The threshold $\theta$ of all cortical excitatory units is updated at the end of each settling phase based on the following 
equations:

@@

\noindent where $\Psi(t)$ is the output of the unit at time t and $\chi=\insval{activity_average_smoothing}$ is a time constant 
controlling the decay of the influence of the past activities. The effect of this scaling mechanism is to bring the average activity of each 
cortical unit closer to the specified target. If the activity in a V1 unit moves away from the target during training, the threshold for activation is thus automatically raised or lowered to bring it closer to the target. Note that an alternative rule with only a single smoothing parameter (rather than
$\xi$ and  $\chi$) could be formulated, but the rule as presented here makes it simple for the modeler to set a desired target activity.

@@

The initial connection field weights are isotropic 2D Gaussians for the lateral excitatory projection and uniformly random within a Gaussian envelope for afferent and lateral inhibitory projections. Specifically, a neuron located at (i, j) will have the following weights in projection $p$:

@@

In the model, as images are presented to the photoreceptors, the cortical afferent connection weights $\omega_{i,j,A}$ from the ON/OFF sheets are adjusted once per iteration (after cortical settling is completed) using a simple Hebbian learning rule. This rule results in connections that reflect correlations between the presynaptic ON/OFF unit activities and the postsynaptic cortical response. Hebbian connection weight adjustment at each iteration is dependent on the presynaptic activity, the postsynaptic response, and the Hebbian learning rate:

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelVariants/drawing.png}
\caption{The four considered intra-cortical connectivity variants. As in figure \ref{fig:model_architecture}, in all panels, the top line represents the excitatory and the bottom line the inhibitory population of neurons. Models (A-C) are using firing rate models of neural units with instantaneous 
translation of inputs into membrane-potential (see section \ref{sec:gcal}). 
(A) Variant 1, assuming only local equally wide excitatory and inhibitory connectivity, and ignoring direct inhibitory to inhibitory interactions. (B) Variant 2, as in A but with added direct inhibitory to inhibitory connections of equal extent as the other excitatory and inhibitory projections. (C) Variant 3, as in A but with added long-range excitatory connections represented based on Buz\'as \etal\,\citep{Buzas2006} as a second wider Gaussian. (D) The same as A but modeled with neural units that take into account membrane time-constant (see section \ref{sec:rate}).
The transmission delay of the inhibitory to excitatory projections in model 4 was varied in the experiments presented in section \ref{sec:SM4}
(figure \ref{fig:figure4}).}
\label{fig:model_variants}
\end{figure}

@@

The architecture of this model (model variant 4; figure \ref{fig:model_architecture} and \ref{fig:model_variants}D) is identical to the GCAL models 
(model variants 1 through 3; figure \ref{fig:model_variants}A-C) with the exception of the equations \ref{eq:vm_model} and \ref{eq:neural_model}, which were replaced with an equation taking into account the membrane time constant:

\begin{equation}
\tau_{z} \frac{\delta\Psi_{i}}{\delta t} = - \Psi_{i} + \sum_{p}\gamma_{p}\sum_{j\in F_{ip}}f(\Psi_{j}(t-\nu_{p}))\omega_{pji} 
\end{equation}
\noindent where $\tau_{z}$ ($z \in \{E,I\}$) is the membrane time constant of excitatory and inhibitory neurons 
($\tau_{E}$ = \insval{membrane_time_constant_E} ms, $\tau_{I}$ = \insval{membrane_time_constant_I} ms), and $\nu_{p}$ is the transmission delay 
for projection $p$ ($\nu_{EE}$ = \insval{td_EE}ms, $\nu_{EI}$ = \insval{td_EI} ms,$\nu_{IE}$ was varied (see section \ref{sec:SM4})).
As a consequence the evolution of the network dynamics has to be simulated at higher resolution. We chose the update step to be \insval{rm_update_step}ms, and we let the activity in the model settle for \insval{rm_settle_time}ms, thus resulting in \insval{rm_settle_steps} settling steps as opposed to the \insval{settle_steps} of the GCAL model. The sub-cortical parametrization of the model is identical to the model variants 1-3 (figure \ref{fig:model_variants}A-C), but because the addition of membrane time constants substantially changes the cortical dynamics
the cortical parameters had to be re-adjusted. Specifically the strength of the thalamo-cortical connections was set to $\gamma_{A}=\insval{w_A}$,
the strength of the lateral excitatory-to-excitatory connections was set to $\gamma_{EE}=\insval{w_EE}$, the strength of the lateral excitatory-to-inhibitory connections was set to $\gamma_{EI}=\insval{w_EI}$, and the strength of the lateral inhibitory-to-excitatory connections was set to $\gamma_{IE}=\insval{w_IE}$.

@@

Model orientation maps are calculated based on the vector average method \citep{CMVC}. We first determine the preferred frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean preferred spatial frequency across all cortical neurons as the
value for the spatial frequency parameter across all subsequent analysis. Next, sinusoidal grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180$\degree$), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

\noindent and thus due to the symmetry of the normal distribution:

@@

\noindent Because convolution of two Normal distributions of variance $\sigma_{i}$ and $\sigma_{e}$ is a normal distribution with variance $\sqrt{\sigma_{i}^2 + \sigma_{e}^2}$ and 
because $\sigma_{i} = \sigma_{e}$ we can simplify to:

@@

\item The change in point 1 will result in a change in the balance of overall excitation and inhibition in the model which is critical 
for successful functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes due to modification 1.

\item Analogously to 2, the changes in 1 change the overall balance between feedforward and lateral contributions to model a cortical neuron's activity, 
which we will compensate by changing the overall strength of the lateral interactions.
